Computer-Assisted Analysis of Emendations of G\"odel's Ontological Proof

### 2014-10-21 Papers/InProgress/LogicAndReligionCongress/Abstract.tex
Explain constant and varying domains

% \begin{abstract}
% The \emph{modal collapse} that afflicts G\"odel's modal ontological 
% argument for God's existence is discussed from the perspective of the 
% modal square of opposition. Furthermore, a computer-assisted verification 
% of the claims that the emendations by Anderson and by Frode are immune to 
% the modal collapse is presented.
% \end{abstract}

%%% ----------------------------------------------------------------------
\maketitle
%%% ----------------------------------------------------------------------
%\tableofcontents
%\section{Introduction}

\vspace*{-5em}

%\section{The Modal Collapse}
\noindent
The axioms in G\"odel's ontological argument 
\cite{GoedelNotes,ScottNotes} (cf. Appendix \ref{apx:Goedel}) entail what is called the \emph{modal collapse}
\cite{Sobel1987,SobelBook2004}: the formula $\varphi \rightarrow \Box
\varphi$, abbreviated as MC, holds for any formula $\varphi$ and not
just for $\exists x. \mathit{God}(x)$ as intended. This fact, which has
recently been confirmed with higher-order automated theorem provers
\cite{C40,J30}, has led to strong criticism of the argument and stimulated attempts to remedy the problem. Hájek
\cite{Hajek_der_Mathematiker_2002,Hajek_Magari_and_others_1996} proposed the use
of cautious instead of full comprehension principles, and Fitting \cite{fitting02:_types_tableaus_god} suggested that greater care is necessary in the semantics of higher-order quantifiers in the presence of modalities. Others, such as
Anderson
\cite{anderson90:_some_emend_of_goedel_ontol_proof,AndersonGettings}
and Bjordal \cite{bjordal99}, proposed slight emendations of
G\"odel's axioms and definitions. They require neither comprehension restrictions nor more complex semantics. Therefore, they are technically simpler to analyze with computer support. We have formalized them using the proof
assistant Isabelle/HOL \cite{Isabelle} together with the automated higher-order
reasoners Leo-II \cite{C26}, Satallax \cite{brown2012satallax}, Metis
\cite{Hurd03first-orderproof}, and Nitpick \cite{Nitpick}.  
Our formalizations\footnote{The
  formalizations are available in the subdirectories \url{Anderson}
  and \url{Bjordal} at
  \url{https://github.com/FormalTheology/GoedelGod/blob/master/Formalizations/Isabelle/}.}
employ the embedding of higher-order modal logic (HOML) in classical
higher-order logic (HOL) as introduced in previous
work \cite{C40,J30,J23}. We explored the effect of different domain conditions
 on the provability of lemmas, theorems and even axioms. This was motivated by a controversy between Hájek and Anderson regarding the redundancy of some axioms in Anderson's emendation. In \emph{constant domain semantics}, the individual domains are the same in all possible worlds. In \emph{varying domain semantics}, the domains may vary from world to world. This variation is technically encoded with the help of an existence relation expressing which individuals actually exist in each world. Quantifiers are then uniformly formalized as \emph{actualist quantifiers} (i.e. guarded by the existence relation). Our main results are summarized in the following sections.

@@

\BWP{Bruno: I do
not yet address T2', L1' and L2' below; we could eventually include those as well?}

For both \textbf{constant domain semantics} and \textbf{varying domain semantics}, the following results hold: 
%
G\"odel's theorems T1, C and T3' can be quickly automated (in logics \K and \KB, respectively); \BWP{3->2 respectively}
%
G\"odel's axioms A4 and A5 are proven redundant (the former one in
  logic \KFourB and the latter one already in \K).
%
Moreover, a trivial countermodel (consisting of two worlds and 
two individuals) to MC generated by Nitpick (for all mentioned logics); this model also shows the consistency of the axioms.

@@

\paragraph{Mixed variant}
%\footnote{Cf. Isabelle/HOL source files \url{Anderson/Anderson_mixed_domain_1/2.thy}} 
(varying domain
  quantifiers are used only in the definitions of essence and NE;
  cf.~Fitting's comments to Anderson in
  \cite{fitting02:_types_tableaus_god}): Also in this setting we
  obtain the same results as above. However, if a varying domain
  quantifier is used only in the definition of NE, then the
  situation changes slightly. Now axiom A5 is no longer provable and
  a countermodel is reported by Nitpick. The remaining results are as before.

@@

% \section{Other Solutions}

% ToDo: Fitting

% \section{Conclusions}

% Various slightly different versions of
% axioms and definitions have been considered by G\"{o}del and by
% several philosophers who commented on his proof
% (cf{sobel2004logic,anderson90:_some_emend_of_goedel_ontol_proof,AndersonGettings,Fitting,Adams,ContemporaryBibliography}).

% In theoretical philosophy, formal logical confrontations with such
% ontological arguments had been so far (mainly) limited to paper
% and pen.  Up to now, the use of computers was prevented, because the
% logics of the available theorem proving systems were not expressive
% enough to formalize the abstract concepts adequately. G{\"o}del's proof
% uses, for example, a complex higher-order modal logic (\HOML) to handle
% concepts such as \emph{possibility} and \emph{necessity} and to support
% quantification over individuals and properties.

% controversies, care with parameters

% ToDo: Leibniz calculemus, Rushby, Zalta \cite{oppenheimera11,rushby13}

% \begin{defn}
% This serves as environment for definitions. Note that the text
% appears not in italics.
% \end{defn}

% \begin{equation}\label{testequation}
% \text{This is a sample equation: } c^2=a^2+b^2
% \end{equation}

% \begin{thm}[Main Theorem]
% In contrast to definitions, theorems appear typeset in italics as
% it has become more or less standard in most textbooks and
% monographs. Equations can be cited using the \verb+\eqref+ command which
% automatically adds brackets: \verb+\eqref{testequation}+ results in \eqref{testequation}.
% \end{thm}

% \begin{proof}
% A special environment is predefined: the \textit{proof} environment. Please use
% \begin{verbatim}\begin{proof}\end{verbatim}
% proof of the statement
% \begin{verbatim}\end{proof}\end{verbatim}
% for typesetting your proofs. The end-of-proof symbol $\Box$ will be added automatically.
% \end{proof}

% There are two known problems with the placement of the end-of-proof sign:

% \begin{enumerate}
%   \item if your proof ends with a\ \ s i n g l e\ \ displayed line, the end-of-proof sign would
% be placed in the line below; if you want to avoid this, write your line in the form
% \begin{verbatim}$$displayed math line \eqno\qedhere$$\end{verbatim}
% which results in

% \begin{proof}
% $$displayed math line \eqno\qedhere$$
% \end{proof}
% \item if your proof ends with an aligned displayed environment, the command
% \verb+\tag*{\qed}+ can be used to place the end-of-proof sign properly:
% \begin{verbatim}
% \begin{align*}
% \alpha&=\beta+\gamma\\
% &=\delta+\epsilon\tag*{\qed}
% \end{align*}
% \end{verbatim}
% results in
% \begin{align*}
% \alpha&=\beta+\gamma\\
% &=\delta+\epsilon\tag*{\qed}
% \end{align*}
% \end{enumerate}
% Please try to avoid using the obsolete \verb+\eqnarray+ environment. This environment has several bugs
% and has been replaced by the more flexible \AmS\ environments \verb+align, split, multline+.

% \begin{rem}
% Additional comments are being typeset without boldfaced entrance
% word as they may be minor important.
% \end{rem}

% \begin{ex}
% For some constructs, even no number is required.
% \end{ex}

% Displayed equations may be numbered like the following one:
% \begin{equation}
% \sqrt{1-\sin^2(x)}=|\cos(x)|.
% \end{equation}

% ------------------------------------------------------------------------

% \subsection*{Acknowledgment}
% Many thanks to ...

@@

% \begin{thebibliography}{1}

% \bibitem{Adams}
% R.M. Adams, `Introductory note to *1970', in {\em {Kurt G\"odel: Collected
%   Works Vol. 3: Unpubl. Essays and Letters}}, Oxford Univ. Press, (1995).

% \bibitem{AndersonGettings}
% A.C. Anderson and M.~Gettings, `G\"odel ontological proof revisited', in {\em
%   {G\"odel'96: Logical Foundations of Mathematics, Computer Science, and
%   Physics: Lecture Notes in Logic 6}},  167--172, {Springer}, (1996).

% \bibitem{anderson90:_some_emend_of_goedel_ontol_proof}
% C.A. Anderson, `Some emendations of {G{\"o}del's} ontological proof', {\em
%   Faith and Philosophy}, {\bf 7}(3), (1990).

% \bibitem{Andrews:gmae72}
% P.B. Andrews, `General models and extensionality', {\em Journal of Symbolic
%   Logic}, {\bf 37}(2),  395--397, (1972).

% \bibitem{andrewsSEP}
% P.B. Andrews, `Church's type theory', in {\em The Stanford Encyclopedia of
%   Philosophy}, ed., E.N. Zalta, spring 2014 edn., (2014).

% \bibitem{C36}
% C.~Benzm{\"u}ller, `{HOL} based universal reasoning', in {\em Handbook of the
%   4th World Congress and School on Universal Logic}, ed., J.Y. Beziau~et al.,
%   pp. 232--233, Rio de Janeiro, Brazil, (2013).

% \bibitem{B5}
% C.~Benzm{\"u}ller and D.~Miller, `Automation of higher-order logic', in {\em
%   Handbook of the History of Logic, Volume 9 --- Logic and Computation},
%   Elsevier, (2014).
% \newblock Forthcoming; preliminary version available at
%   {http://christoph-benzmueller.de/papers/B5.pdf}.

% \bibitem{C34}
% C.~Benzm{\"u}ller, J.~Otten, and Th. Raths, `Implementing and evaluating
%   provers for first-order modal logics', in {\em Proc. of the 20th European
%   Conference on Artificial Intelligence (ECAI)}, pp. 163--168, (2012).

% \bibitem{B9}
% C.~Benzm{\"u}ller and L.C. Paulson, `Exploring properties of normal multimodal
%   logics in simple type theory with {LEO-II}', in {\em {Festschrift in Honor of
%   {Peter B. Andrews} on His 70th Birthday}}, ed., C.~Benzm{\"u}ller~et al.,
%   386--406, College Publications, (2008).

% \bibitem{J23}
% C.~Benzm{\"u}ller and L.C. Paulson, `Quantified multimodal logics in simple
%   type theory', {\em Logica Universalis}, {\bf 7}(1),  7--20, (2013).

% \bibitem{LEO-II}
% C.~Benzm{\"u}ller, F.~Theiss, L.~Paulson, and A.~Fietzke, `{LEO-II} - a
%   cooperative automatic theorem prover for higher-order logic', in {\em
%   Proc.~of IJCAR 2008}, number 5195 in LNAI, pp. 162--170. Springer, (2008).

% \bibitem{J30}
% C.~Benzm{\"u}ller and B.~Woltzenlogel-Paleo, `Formalization, mechanization and
%   automation of {G{\"o}del's} proof of {God's} existence', {\em
%   arXiv:1308.4526}, (2013).

% \bibitem{J28}
% C.~Benzm\"uller and B.~Woltzenlogel-Paleo, `{G{\"o}del's God in Isabelle/HOL}',
%   {\em Archive of Formal Proofs}, (2013).

% \bibitem{W50}
% C.~Benzm\"uller and B.~Woltzenlogel-Paleo, `G\"odel's {God} on the computer',
%   in {\em Proceedings of the 10th International Workshop on the Implementation
%   of Logics}, EPiC Series. EasyChair, (2013).
% \newblock Invited abstract.

% \bibitem{Coq}
% Y.~Bertot and P.~Casteran, {\em {Interactive Theorem Proving and Program
%   Development}}, Springer, 2004.

% \bibitem{Nitpick}
% J.C. Blanchette and T.~Nipkow, `Nitpick: A counterexample generator for
%   higher-order logic based on a relational model finder', in {\em Proc. of ITP
%   2010}, number 6172 in LNCS, pp. 131--146. Springer, (2010).

% \bibitem{Satallax}
% C.E. Brown, `Satallax: An automated higher-order prover', in {\em Proc. of
%   IJCAR 2012}, number 7364 in LNAI, pp. 111 -- 117. Springer, (2012).

% \bibitem{ContemporaryBibliography}
% R.~Corazzon.
% \newblock Contemporary~bibliography~on~ontological~arguments: {\scriptsize
%   \url{http://www.ontology.co/biblio/ontological-proof-contemporary-biblio.htm}}.

% \bibitem{Fitting}
% M.~Fitting, {\em Types, Tableaux and G\"odel's God}, Kluwer, 2002.

% \bibitem{fitting98}
% M.~Fitting and R.L. Mendelsohn, {\em First-Order Modal Logic}, volume 277 of
%   {\em Synthese Library}, Kluwer, 1998.

% \bibitem{Gallin75}
% D.~Gallin, {\em Intensional and Higher-Order Modal Logic}, North-Holland, 1975.

% \bibitem{garbacz12:_prover_simpl_expal_away}
% P.~Garbacz, `{PROVER9's} simplifications explained away', {\em Australasian
%   Journal of Philosophy}, {\bf 90}(3),  585--592, (2012).

% \bibitem{GoedelNotes}
% K.~G\"odel, {\em Appx.A: Notes in Kurt G\"odel's Hand},  144--145.
% \newblock In  \cite{sobel2004logic}, 2004.

% \bibitem{Henkin50}
% L.~Henkin, `Completeness in the theory of types', {\em Journal of Symbolic
%   Logic}, {\bf 15}(2),  81--91, (1950).

% \bibitem{homl}
% R.~Muskens, `{Higher Order Modal Logic}', in {\em Handbook of Modal Logic},
%   ed., P~Blackburn~et al.,  621--653, Elsevier, Dordrecht, (2006).

% \bibitem{Isabelle}
% T.~Nipkow, L.C. Paulson, and M.~Wenzel, {\em {Isabelle/HOL: A Proof Assistant
%   for Higher-Order Logic}}, number 2283 in LNCS, Springer, 2002.

% \bibitem{oppenheimera11}
% P.E. Oppenheimer and E.N. Zalta, `A computationally-discovered simplification
%   of the ontological argument', {\em Australasian Journal of Philosophy}, {\bf
%   89}(2),  333--349, (2011).

% \bibitem{rushby13}
% J.~Rushby, `The ontological argument in {PVS}', in {\em Proc.~of CAV Workshop
%   ``Fun With Formal Methods''}, St. Petersburg, Russia,, (2013).

% \bibitem{Schulz:AICOM-2002}
% S.~Schulz, `E -- a brainiac theorem prover', {\em {AI Communications}}, {\bf
%   15}(2),  111--126, (2002).

% \bibitem{ScottNotes}
% D.~Scott, {\em Appx.B: Notes in Dana Scott's Hand},  145--146.
% \newblock In  \cite{sobel2004logic}, 2004.

% \bibitem{sobel2004logic}
% J.H. Sobel, {\em Logic and Theism: Arguments for and Against Beliefs in God},
%   Cambridge U. Press, 2004.

% \bibitem{sutcliffe2009tptp}
% G.~Sutcliffe, `The {TPTP} problem library and associated infrastructure', {\em
%   Journal of Automated Reasoning}, {\bf 43}(4),  337--362, (2009).

% \bibitem{J22}
% G.~Sutcliffe and C.~Benzm{\"u}ller, `Automated reasoning in higher-order logic
%   using the {TPTP THF} infrastructure.', {\em Journal of Formalized Reasoning},
%   {\bf 3}(1),  1--27, (2010).

% \bibitem{J29}
% B.~Woltzenlogel-Paleo and C.~Benzm{\"u}ller, `Automated verification and
%   reconstruction of {G\"odel's} proof of {God's} existence', {\em OCG J.},
%   (2013).

%\end{thebibliography}

\newpage
\begin{appendix}
\noindent @Bruno: I think we need more compact presentation of the proofs.
% \begin{figure}[t]
% \noindent \framebox[\columnwidth][r]{
% \begin{minipage}{.94\columnwidth}\small
% \begin{itemize}
% \item[MC] Everything that is the case is so necessarily:
%   $$\hol{\allq \phi [\phi \imp \nec \phi]}$$ 
% \item[MC'] Everything that is possible is necessary:
%   $$\hol{\allq \phi [\pos \phi \imp \nec \phi]}$$ 
% \item[MC''] Modalities collapse:
%   $$\hol{\allq \phi [\phi \biimp \pos \phi \biimp \nec \phi]}$$
% \end{itemize}
% \end{minipage}
% } \vskip-.5em
% \caption{Modal Collapse \cite{todo:Sobel1987,SobelBook}.\label{fig:collapse}} 
% \end{figure}

>>>>

% \begin{abstract}
% \end{abstract}

%%% ----------------------------------------------------------------------
\maketitle
%%% ----------------------------------------------------------------------
%\tableofcontents

\noindent The axioms in G\"odel's ontological argument
\cite{GoedelNotes,ScottNotes} (cf. Appendix \ref{apx:Goedel}) entail
what is called \emph{modal collapse}
\cite{Sobel1987,SobelBook2004}: the formula $\varphi \rightarrow \Box
\varphi$, abbreviated as MC, holds for any formula $\varphi$ and not
just for $\exists x. \mathit{God}(x)$ as intended. This fact, which
has recently been confirmed with higher-order automated theorem
provers \cite{C40,J30}, has led to strong criticism of the argument
and stimulated attempts to remedy the problem. Hájek
\cite{Hajek_der_Mathematiker_2002,Hajek_Magari_and_others_1996}
proposed the use of cautious instead of full comprehension principles,
and Fitting \cite{fitting02:_types_tableaus_god} suggested that
greater care is necessary in the semantics of higher-order quantifiers
in the presence of modalities. Others, such as Anderson
\cite{anderson90:_some_emend_of_goedel_ontol_proof,AndersonGettings}
and Bjordal \cite{bjordal99}, proposed slight emendations of G\"odel's
axioms and definitions. They require neither comprehension
restrictions nor more complex semantics. Therefore, they are
technically simpler to analyze with computer support. We have
formalized them using the proof assistant Isabelle/HOL \cite{Isabelle}
together with the automated higher-order reasoners Leo-II \cite{C26},
Satallax \cite{brown2012satallax}, Metis 
\cite{Hurd03first-orderproof}, and Nitpick \cite{Nitpick}.   Our
formalizations\footnote{The   formalizations are available in the
subdirectories \url{Anderson}   and \url{Bjordal} at   \url{https://gi
thub.com/FormalTheology/GoedelGod/blob/master/Formalizations/Isabelle/
}.} employ the embedding of higher-order modal logic (HOML) in
classical higher-order logic (HOL) as introduced in previous work
\cite{C40,J30,J23}. We explored the effect of different domain
conditions  on the provability of lemmas, theorems and even axioms.
This was motivated by a controversy between Hájek and Anderson
regarding the redundancy of some axioms in Anderson's emendation. In
\emph{constant domain semantics}, the individual domains are the same
in all possible worlds. In \emph{varying domain semantics}, the
domains may vary from world to world. This variation is technically
encoded with the help of an existence relation expressing which
individuals actually exist in each world. Quantifiers are then
uniformly formalized as \emph{actualistic quantifiers} (i.e. guarded by
the existence relation). Our main results are summarized in the
following sections.

@@

%\BWP{Bruno: I do not yet address T2', L1' and L2' below; we could eventually include those as well?}

For both \textbf{constant domain semantics} and \textbf{varying domain semantics}, the following results hold: 
%
the theorems T1, C and T3' can be quickly automated (in logics \K, \K and \KB, respectively);
%
the axioms A4 and A5 are proven redundant (the former in
logic \KFourB and the latter already in \K);
%
a trivial countermodel (consisting of two worlds and 
two individuals) to MC generated by Nitpick (for all mentioned logics), which also shows the consistency of the axioms.

The redundancy of A4 and A5 is particularly controversial. Magari \cite{TODO} claimed that the redundancy occurs already with Gödel's original axioms and definitions. Hájek \cite{Hajek 1996 TODO} investigated this further and claimed that Magari's claim is invalid, but true under the assumption of an additional axiom (PEP). Moreover, Hájek \cite{Hajek 1996 TODO} claimed that the redundancy occurs for Anderson's emended axioms and definitions \cite{Anderson 1990}. Anderson and Gettings \cite{AG1996, footnote 1, page 1} rebutted Hájek's claim, arguing that it holds only under constant domain semantics, while the emended argument by Anderson ought to be taken under Cocchiarella's semantics \cite{Cocchiarella} (a particular kind of varying domain semantics). Hájek \cite{Hajek 2002, page 7} acknowledges this rebuttal, and apparently accepts it, as evidenced by his use of A4 and A5 in his new small emendation (replacing A:A1 and A2 by a new axiom H:A12) of Anderson's variant with varying domains \cite{Hajek 2002, section 4}. Nevertheless, he does show yet another emended version where A4 and A5 are redundant, but A3 is replaced by a stronger axiom additionally stating that the property of actual existence is positive \cite{Hajek 2002, section 5}. Our results show that Hájek was originally right, under both constant and varying domain semantics with no need to strengthen A3 (though this need might exist when H:A12 is used instead of A:A1 and A2). It should be noted, however, that \cite{Anderson, footnote 14} vaguely remarks that only the quantifiers in T3' and in A:D2  need to be interpreted as actualistic quantifiers, while other may be taken as possibilistic quantifiers. We have inv  mixed variant.

@@

\paragraph{Mixed variant}
(varying domain
  quantifiers are used only in the definitions of essence and NE;
  cf.~Fitting's comments to Anderson in
  \cite{fitting02:_types_tableaus_god}): Also in this setting we
  obtain the same results as above. However, if a varying domain
  quantifier is used only in the definition of NE, then the
  situation changes slightly. Now axiom A5 is no longer provable and
  a countermodel is reported by Nitpick. The remaining results are as before.

@@

@@

\newpage
\begin{appendix}
\noindent

\BWP{I think we need more compact presentation of the proofs.}
